---
title: "Report"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
params:
  sim_params: NA
  tbl_T: NA
  tbl_NT: NA
  hcr_plot: NA
  harv_plot: NA
  NOE_plot: NA
  NOS_plot: NA
  spawners_plot: NA
  NOE_ratios_plot: NA
  RMRS_plot: NA
  hatch_plot: NA
  h_surplus: NA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.width = 7)
```


### Parameters used in sumulations

  *Harv_mod* is the choice for model of allowed vs. realized harvest. The "log-log" model assumes that the average harvest will be less than the allowed, the "broken-stick" model assumes that in-river harvest is not responsive (due to management error) at lower allowed harvest rates, and the "average" model assumes that harvest will be equal to allowed on average
  
  *HOS_mod* indicates wheather hatchery-origin spawners were included. "HOE" indicates that they were included and "zero" indicated that they were not
  
  *URR* is the proportion of unmarked fish captured in the non-treaty fisheries that are released.
  
  *IE* is a scalar for implementation error in harvest (i.e., due to forecast and managment error). The defaults are 20% for treaty and 0.15% for non-treaty with the default model, and 21% for treaty and .18% for non-treaty with the broken hockey stick model. The scalar is multiplied by those default values.
  
```{r}
knitr::kable(params$sim_params)
```


### simulation results

```{r, fig.cap=" Harvest rates for different sectors and all sectors combined under alternative harvest control rules. The denominator in the rates shown is the river mouth run size, which is different from what is used to calculate allowable impacts in the the current Agreement. River mouth run size plus PFMC non-treaty AEQ mortalities is used as the denominator in the current Agreement. The plots assume that PFMC AEQ non-treary mortality is at average rates.",fig.width = 7,fig.height = 8}
(params$hcr_plot)
```


```{r, fig.cap="Box plots of annual harvest. The whiskers of the boxplot span 95% quantile intervals, the box spans the 50% quantile interval, and the midline is the median. Quantiles were calculated across years for individual simulations, and then averaged across simulations.  Therefore, the lower end of the bars represent the average across simulations of the 2 years (out of 25) with the lowest harvest. I summarized the results in this way to show what harvest projections were in small run size, average, and large run size years."}
(params$harv_plot)
```

```{r, fig.cap="Box plots of  natural-origin escapement, which includes spawning, hatchery broodstock, and hatchery surplus. The whiskers of the boxplot span 95% quantile intervals, the box spans the 50% quantile interval, and the midline is the median. Quantiles were calculated across years for individual simulations, and then averaged across simulations.  Therefore, the lower end of the bars represent the average across simulations of the 2 years (out of 25) with the lowest escapement I summarized the results in this way to show what escapement projections were in small run size, average, and large run size years.",fig.width = 7,fig.height=8}
print(params$NOE_plot)
```

```{r, fig.cap="Quantile of natural-origin spawners across years of simulations."}
print(params$NOS_plot)
```

```{r, fig.cap="Quantile of total spawners across years of simulations."}
print(params$spawners_plot)
```

```{r, fig.cap="Ratio of escapement to unfished escapement. Bars in the left panel show the ratio of geometric mean escapement between a given harvest control rule and a no-terminal-fishing control rule."}
print(params$NOE_ratios_plot)
```

```{r, fig.cap= "Boxplots show quantiles of River Mouth Run sizes. Note, the river mouth run size can increase with some harvest due to overcompensation (i.e., decreasing productivty at higher spawner abundances) in the Ricker model." }
print(params$RMRS_plot)
```

```{r, fig.cap=" Proportion of hatchery origin spawners (pHOS), proporiton of natural origin broodstock (pNOB) and proportionate natural influence (PNI= pNOB/(pNOB+pHOS)).Quantiles were calculated across years for individual simulations, and then averaged across simulations.The whiskers of the boxplot span 95% quantile intervals, the box spans the 50% quantile interval, and the midline is the median.",fig.width = 7,fig.height=8}
print(params$hatch_plot)
```

```{r, fig.cap="Quantiles of number of hatchery origin returns that were not harvested nor needed for broodstock."}
print(params$h_surplus)
```

## Harvest control rule definitions

The current harvest control rule specifies fixed rates numbers or rates in lower abundance tiers, and in higher abundance tiers the allowable catch is calculated as:
              $$(scalar * run size - offset) * share$$


- *tiers* = the run size below which the rates in the row are applied.
- *rate* = if a value is entered in the rate column, that is the harvest rate for the tier. If the value is greater than 1, it is assumed to be a total allowable catch rather than a rate.
- *scalar* = if the allowable catch is a function of the run size, as in the two highest tiers of the current rule, this number is multiplied by the run size.
- *offset* = if the allowable catch is a function of the run size, as in the two highest tiers of the current rule, this number is subtracted from the scaled run size.
- *share* = if the allowable catch is a function of the run size, as in the two highest tiers of the current rule, this number is multipled by the scales run size less the offset.


The abundance index for the current harvest control rule is the total river mouth run size plus calendar year PFMC AEQ mortalities. If the value in the index column in the tables below is "wild" then that harvest control rule was based on an index representing wild abundance at the river mouth.


THe current harvest control rules includes PFMC AEQ morts in the non-treaty share at run size >29,000. There is a column in the tables below that specifies teh run size above which PFMC AEQ mortalities were included in the Non-treaty share. 


### Non-treaty

```{r}
knitr::kable(params$tbl_NT,caption = "Non-treaty harvest control rule definitions")
```



### Treaty

```{r}
knitr::kable(params$tbl_T,caption = "Treaty harvest control rule definitions")
```


